---
title: "Te Tutorial"
author: "Lauren Buckley"
date: "8/05/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

energy balance

<!-- Gates example -->

```{r}
##Load TrenchR package
library(TrenchR)
```

##Operative environmental temperatures

##Energy balance and components
Calculations of operative environmental temperatures are based in balancing energy input and output to organisms.  The basic equation [Gates 1980] is
$$Qnet = Qabs - Qemit + Qconv + Qcond + Qmet - Qevap$$
where $Qnet$ is the net energy exhange with the environment (W), $Qabs$ is the solar and thermal radiation absorbed (W), $Qemit$ is the thermal radiation emitted (W), $Qconv$ is energy exchange due to convection (W), $Qcond$ is energy exchange due to conduction (W),
$Qmet$ is the energy emitted due to metabolism (W), and $Qevap$ is the energy emitted due to evaporative water loss (W).

The energy balance is available in TrenchR as follows:
```{r}
##Qnet_Gates(Qabs, Qconv, Qcond, Qmet, Qevap)
##Qnet_Gates(Qabs=500, Qemit=10, Qconv=100, Qcond=100, Qmet=10, Qevap=5)
```

#Qabs

#Qemit

#Qconv

#Qcond

#Qmet

#Qevap

#Function Tb_Gates
The following function can be used to estimate Te,
```{r}
Tb_Gates(As=1, d=0.001, psa_dir=0.6, psa_ref=0.4, psa_air=0.6, psa_g=0.2, Ts=303, Ta=310, Rabs=800, emissivity=0.95, h_L=10, ef=1.3, K=0.5)
```

##Alternative energy balance models and functions
Tb_CN
Tb_Fei

##Specialized energy balance models and functions
Tb_lizard
Tb_butterfly
Tb_grasshopper
Tb_salamander humid
