---
title: "Allometries and conversions"
author: "Lauren Buckley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Allometry}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
##Load TrenchR package
library(TrenchR)
```

#Allometries
We provide tools for estimating additional dimensions of organisms from measured dimensions. The function `mass_from_length()` encompasses empirically-derived relationships to estimate mass (mostly) as a power law function of mass for six taxonomic classifications. The relationships are derived by relating length to mass for specimens of numerous species and avaialble as the following function: 
```{r}
mass_from_length(l=0.04,taxa="insect")
```

Surface area is needed for several aspects of biophysical models but is rarely meausured. We provide functions to estimate surface area ($m^3$) from length, mass, or volume for a variety of taxa. The function `sa_from_mass()` encompasses empirically-derived relationship to estimate surface area as a power law function of mass for lizards, frogs, and insects.  We also provide empirical relationships estimating surface area as power law function of volume. The functions were developed by Mitchell [1976] for use in approximating convective heat transfer. For the same study, we additionally provide a function to estimate volume from length for several taxa or approximating the organism as a sphere. Finally, we provide a function for estimating surface area from length by approximating the animal's body as a rotational ellipsoid. The allometric functions are available in R as follows:
```{r}
sa_from_mass(2,"insect")
sa_from_volume(V=0.001,taxa="lizard")
volume_from_length(l=0.05,taxa="lizard")
sa_from_length(l=0.04)
```

Estimating the amount of solar radiation that an animal absorbs requires knowledge of the silhouette area, the propotion of an animal's surface area that is projected onto a horizontal surface. Silhouette area is a function of the sun's angle. We offer empirically-derived relationships to predict the proportional silhouette area for frogs, lizards, and grasshoppers as a polynomial function of zenith angle (`prop_silhouette_area()`).  We additionally offer a function (`prop_silhouette_area_shapes()`) to estimate silhoutte area by approximating the animal as a sphere or cylinder.  The R functions are parameterized as follows:
```{r}
prop_silhouette_area(60, taxa= "frog")
prop_silhouette_area_shapes(shape="spheroid", theta=60, h=0.01, d=0.001)
```

#Conversions
We offer convenience functions to convert among units for angles and temperatures:
```{r}
degree_to_radian(47.608)
radian_to_degree(0.831)
fahrenheit_to_celsius(85)
fahrenheit_to_kelvin(85)
kelvin_to_celsius(270)
```


