---
title: "Estimating microclimates"
author: "Lauren Buckley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
##Load TrenchR package
library(TrenchR)
```

# Radiation
## Solar angles and daylength
Estimates of sun position and day length are often estimated based on calendar date. We provide a function `day_of_year()` that facilitates translating a date into a calendar date. We next present functions to calculate sun angles.  We use the calendar date (J) to estimate solar declination (radian), which is the angular distance of the sun north or south of the earth’s equator, as follows (`dec_angle()`):
$$ \delta= \arcsin[0.39795 \cos(0.21631+ 2 \arctan (0.967 \tan [0.0086(-186+J)]))].$$

Zenith angle ($\psi$, degrees), the location of the sun as an angle measured from vertical, can be estimated as:
$$\cos \psi = \sin( \delta) \sin( \phi) + \cos(\delta)\cos(\phi)\cos(\pi/12*(h-h_0)),$$
where J is calendar date, $\phi$ is latitude (radians), h is hour, and $h_0$ is the time of solar noon. The time of solar noon (`solar_noon()`) can be estimated as $h_0= 12 - LC -ET$, where LC in the longitude correction (+4 minutes for each degree east of the standard meridian) and ET is the equation of time.  The ET is estimated as: 
$$ ET= \frac{-104.7\sin (f)+596.2\sin (2f)+4.3\sin (3f)-12.7\sin (4f)-429.3\cos (f)-2.0\cos (2f)+19.3\cos (3f)}{3600}, $$
where $f=279.575 + 0.9856 J$, in degrees. 

The azimuth angle, the angle (in degrees) from which the sunlight is coming measured from true north or south measured in the horizontal plane. The azimuth angle ($AZ$ in degrees) is estimated with respect to due south, increasing in the counter clockwise directi such that $90^\circ$ is east. $AZ$ can be calculated as a function of declination angle $\delta$, zenith angle $\psi$ and latitude $\phi$ as follows:
$$\cos AZ= \frac{-(\sin \delta - \cos \psi \sin \phi)}{\cos \phi \sin \psi}.$$

These sun angles can be used to solve for daylength. The equation for half daylength $h_s$, which is the time (in degrees) from sunrise to solar noon, is used as the basis for our `daylength()` function:
$$ h_s= \cos^{-1}\left( \frac{\cos \psi-sin \phi \sin \delta}{\cos \phi \cos \delta} \right).$$

We provide these R functions as follows: 
```{r}
day_of_year("2017-04-22", format= "%Y-%m-%d")
dec_angle(112)
zenith_angle(doy=112, lat=47.61, lon=-122.33, hour=12)
solar_noon(lon=-122.335, doy=112)
azimuth_angle(doy=112, lat=47.61, lon=-122.33, hour=12)
daylength(lat=47.61, doy=112)
```

## Radiation components
We provide functions to estimate solar (shortwave) radiation by discounting incoming solar radiation due to scattering as the radiation travels through the atmosphere. The primary function `estimate_radiation()` returns estimates of the direct, diffuse, and reflected components of solar radiation ($W m^{-2}$). The radiation streams are the direct irradiance on a surface perpendicular to the beam $S_p$, the diffuse sky irradiance on a horizontal plane $S_d$ (W m−2), and the reflected radiation from the ground $S_r$. The sum of these three streams represents the total irradiance of a horizontal surface $S_t$.

Calculation of these flux densities requires the introduction of several additional quantities. The atmospheric transmissivity, $\tau$, ranges between 0.6 and 0.7 for typical clear sky conditions (Gates 1980). The solar constant, $S_{p0}$, indicates extraterrestrial flux density to be 1,360 $W m^{−2}$. Optical air mass number, $m_a$, is the ratio of slant path length through the atmosphere to zenith path length and is a function of atmospheric pressure: $m_a=p_a/(101.3\cos(\psi))$, where $p_a$ is air pressure (kPa) and $\psi$ is zenith angle (see above)$. The function `airpressure_from_elev()` estimates air pressure as $p_a (kPa)= 101.3 \exp \frac{-E}{8200}$, where $E$ is the elevation in meters above sea level. 

Direct irradiance is a function of the distance a solar beam travels through the atmosphere; the transmittance of the atmosphere, $\tau$; and the incident flux density, $\S_{p0}$: $S_p=S_{p0}\tau^{m_a}$. We approximate diffuse radiation using an empirical relation [Liu and Jordan 1960], $S_d=0.3(1-\tau^{m_a}) S_{p0}\cos\psi$.Finally, reflected radiation is the product of albedo, $\rho_S$ and the sum of direct and diffuse radiation: $S_r= \rho_S (S_p+S_d)$.

We additionally provide a function (`monthly_solar_radiation()`) implementing an algorithm to estimate average monthly solar radiation ($W m^{-2}$ per day) using basic topographic and climatic information as input.

We provide a function `partition_solar_radiation()` to estimate the diffuse fraction $k_d$ that can be used to partition measures of solar radiation into direct and diffuse components.  The partitioning is based on empirically-derived relations estimated as a function of the clearness index $k_t$, which is the ratio of the global solar radiation measured at the surface to the total solar radiation at the top of the atmosphere. The relations are compiled in Wong and Chow [2001, Applied Energy 69:1991-224].

The functions are available in the TrenchR package as follows:
```{r}
airpressure_from_elev(1500)
estimate_radiation(doy=112, psi=1, tau=0.6, elev=1500, rho=0.7)
#monthly_solar_radiation(lat=47.61,lon=-122.33,doy=112,elev=1500,Temp=15,Hr=50,P=50)
#partition_solar_radiation(500, 0.5)
```

# Temperature and windspeed profiles
Temperatures and windspeeds collected at ~2m can differ drametically from conditions near the ground where most organisms live. Turbulent air flow influenced by the ground surface roughness sets up temperature and windspeed profiles.  These profiles can be estimated to scale temperatures and widspeeds to organism height. We scale across height by estimating a wind speed profile using the relationship:
$$u(z)= \frac{u^*}{0.41}ln \frac{z-d}{z_m} $$
where u(z) is the wind speed (m/s) at height z (m), u* is the friction velocity (m/s), d is the zero plane displacement (m, the height at which wind speed is zero), and the constant 0.41 is the von Karman constant. 

We provide a function to use this relationship to estimate surface roughness (`surface_roughness()`) from wind speed measurements at a vector of heights.  First, the zero-plane displacement is estimated as the intercept of the relationship between the wind velocities and the log of measurement height.  Surface roughness $z_0$ can then be estimated by repeating the regression accounting for the zeroplane displacement (by substracting the zero-plane displacement from the measurement heights). $z_0$ is estimated as the exponential of the ration of the intercept and slope of the regression relationship.

We use the following relationship to estimate air temperature and wind speed at height z under neutral conditions (no buoyancy associated with differential density):
$$\frac{u_z}{u_r}=\frac{ln(z/z_0+1)}{ln(z_r/z_0+1)}=\frac{T_z-T_0}{T_r-T_0} $$
where uz and ur are the wind speeds (m/s) at height z (m) and reference height r (m), respectively.  Tz, Tr, and T0 are air and soil temperatures at height z and r and soil surface temperature, respectively. We provide this relationship in the `wind_speed_profile_neutral()` function to estimate windspeed at a specified height. We additionally provide this relationship in the `air_temp_profile_neutral()` function to estimate temperature at a specified height.

The functions are available as:
```{r}
surface_roughness(u_r=c(0.01,0.025,0.05,0.1,0.2), zr=c(0.05,0.25,0.5,0.75,1))
wind_speed_profile_neutral(u_r=0.1, zr=0.1, z0=0.2, z=0.15)
air_temp_profile_neutral(T_r=20, zr=0.1, z0=0.2, z=0.15, T_s=25)
```

We additionally provide functions for estimating wind speed and temperature profiles for the case of free convection using the relationships in NicheMapR as described by Kearney and Porter [2016].  The function `wind_speed_profile()` estiamtes wind speed at height $z$ as $u_z=2.5u^*\ln(z/z_0+1)$ where the friction velocity $u^*$ is estimated as $u^*=\frac{0.41u_r}{\ln(z/z_0+1)}$ and 0.41 is the von Karman constant.

Temperature at a given height is estimated in the function `air_temp_profile()` by first estimating the temperature at roughness height:
$$T_{z_0}= \frac{T_r S_{t_b}-T_0 S_{t_s}}{S_{t_b}-S_{t_s}}. $$
where $S_{t_s}$ is the sublayer Stanton Number $S_{t_s}=0.62/ (\frac{z_0u^*}{12})^{0.45}$ and $S_{t_b}$ is the bulk Stanton Number $S_{t_b}=0.64/ (\frac{z}{z_0}+1)$. The Stanton numerbers are dimensionless numbers that measures the ratio of the heat transfered in and the thermal capacity. The temperature at height $z$ can then be estimated as:
$$T_z= T_{z_0}+(T_r-T_{z_0})ln(\frac{z}{z_0}+1) .$$
The functions are available in TrenchR as:
```{r}
wind_speed_profile(u_r=0.1, zr=0.1, z0=0.2, z=0.15)
air_temp_profile(T_r=20, u_r=0.1, zr=0.1, z0=0.2, z=0.15, T_s=25)
```

# Diurnal variation in temperature and radiation
We provide three forms of equations to estimate diurnal temperature radiation: a sine model, a sine exponential model, and a sine sqrt model. The models assume that maximum temperature $T_x$ will occur sometime between sunrise $t_r$ and sunset $t_s$ and that the minimum temperature $T_n$ will occur in the early morning. 

The sine model (`diurnal_temp_variation_sine()`) estimates temperature t hour $h$ as a function of minimum and maximum temperature $$T(h)= T_x -T_n(1-\gamma) $$, where $\gamma=0.44-0.46 \sin (0.9+WH_r)+0.11 \sin(0.9+2WH_r)$ and $W=\pi/12$

For the sine exponential model (`diurnal_temp_variation_sineexp()`), hourly temeprature is estimated as a function of three empirically determined parameters: $\alpha$ is the time difference between $h_x$ and midday; $\Beta$ is the time difference between $h_n$ and sunrise; and $\gamma$ is a decay parameter that determines the rate of temperature change from sunset to $h'_n$ of the next day. 

The time of minimum temperature can be calculated as $h_x=1/2(t_r+t_s)+\alpha$. The time of maximum temperature can be calculates as $t_n=t_r+\beta$. The model estimates temperature at time $h$ as follows: $$  T(h)=\left \{
                \begin{array}{ll}
                  T_n+(T_x-T_n)\sin \frac{\pi(h-h_r-\beta)}{l +2(\alpha-\beta)}, h_n \leq h \leq h_s\\
                  T_n'+[T(h_s)-T'_n]\exp \frac{\gamma(h-h_s)}{24-l+\beta}, h_s \leq h \leq h'_n 
                \end{array}
              \right. $$
where $h'_n$ is the time of the minimum temeprature, $T'_n$, for the next day. Daylength $l$ is calculated as $h_s-h_r$. 

The sine square root model (`diurnal_temp_variation_sinesqrt()`) estimates temperature assuming the time of maximum temperature $t_x$ occurs 4 hours before sunset ($t_x=t_s -4$):
$$  T(h)=\left \{
                \begin{array}{ll}
                  T_n+\alpha \pi /2 \left \{ \frac{h-h_r}{h_x-h_r} \right \}, h_r < h \leq h_x\\
                  T_r+R \sin [\pi/2+(\frac{h-h_x}{4})\pi/2], h_x < h < h_s\\ 
                  T_r+b \sqrt{h-h_s}, h_s < h \leq h_p
                \end{array}
              \right. $$
where $\alpha=T_x-T_n$,$R=T_x-T_s$, and $b=\frac{T_p-T_s}{\sqrt{h_p - h_s}}$. Sunset temperature is estimated as $T_s=T_x-c(T_x-T_p)$ where $c$ is an empicially estimated parameter ($c=0.39$).

We provide the functions as follows:
```{r}
diurnal_temp_variation_sine(T_max=30, T_min=10, t=11)
diurnal_temp_variation_sineexp(T_max=40, T_min=10, t=11, t_r=6, t_s=18, alpha=2.59, beta= 1.55, gamma=2.2)
diurnal_temp_variation_sinesqrt( t=8, tr=6, ts=18, T_max=30, T_min=20, T_minp=22)
```

We adapt equations from Tham et al. [2010] to estimate hourly solar radiation ($W m^{-2}$ per hour) as a function of daily global solar radiation (in $W m^{-2}$ per day). The function `diurnal_radiation_variation()` partitions daily solar radiation $S$. The function estimates the ratio of hourly to daily global irradiation,$S_G$, as follows:
$$S_G= \frac{\pi}{24}(a'+b'\cos \omega)\frac{\cos \omega - \cos \omega_s}{\sin \omega_s - \omega_s \cos \omega_s}, $$
where $a'= 0.409+0.5016\sin(\omega_s-1.047)$ and $b'=0.6609-0.4767\sin(\omega_s-1.047)$. The function utilizes two angles in radians: $\omega$ is the hour angle of the sun and $\omega_s$ is the hour angle at sunset.  The function calculates these angles as a function of calendar date $J$ and location (latitude and longitude):
```{r}
diurnal_radiation_variation(doy=112, solrad=500, hour=12, lon=-122.33, lat=47.61)
```

# Degree days
Many organismal processes and rates are temperature dependent. Accounting for temperature dependence allows translating calendar time into a more biologically relevant concept known as degree days. Degree days indicates the accumulated heat available for processes such as development and is estimated as the accumulated product of time and temperature. For the frequent application of estimating development, degree days are calculated within physiological bounds known as the lower and upper developmental temperatures, LDT and UDT, respectively.  We provide four approaches for estimating degree days that differ in the complexity with which the aggregate heat: "single.sine"","double.sine", "single.triangulation" or "double.triangulation". We implement the methods detailed here: http://ipm.ucanr.edu/WEATHER/ddfigindex.html.  The methods account for whether the temeprature time series is intercepted by the LDT and / or the UDT.  Briefly, we present the equations for each method for the case of the temperature time series being intercepted by the lower threshold. For the single-sine method, degree days $D$ can be estimated as
$$D= \frac{1}{\pi} \left[ \left( \frac{T_x+T_n}{2}-LDT \right) +\alpha\cos(\theta_1) \right] \\
\theta_1= \sin^{-1}\left[ 1/ \alpha \left( LDT- \frac{T_x+T_n}{2} \right) \right]$$

For the double-sine method, degree days $D$ can be estimated as
$$D= \frac{1}{2\pi} \left[ \left( \frac{T_x+T_n}{2}-LDT \right)(\pi/2- \theta_1 ) +\alpha\cos(\theta_1) \right] \\
\theta_1= \sin^{-1}\left[ 1/ \alpha \left( LDT- \frac{T_x+T_n}{2} \right) \right].$$

For the single triangulation method, degree days $D$ can be estimated as
$$D= \frac{(T_x-LDT)^2}{2(T_x-T_n)} .$$
For the double triangulation method, degree days $D$ can be estimated as
$$D= \frac{(T_x-LDT)^2}{4(T_x-T_n)}. $$

The methods are available in the following R function:
```{r}
degree_days(T_min=7, T_max=14, LDT=12, UDT=33, method="single.sine")
degree_days(T_min=7, T_max=14, LDT=12, UDT=33, method="single.triangulation")

```

# Soil temperatures
We implemented a microclimate model with air temperature and radiation as input using finite-difference methods to solve heat balance equations describing soil temperatures at the surface and specified depths. Our approach follows Porter et al. [1973] and is similar to that described for the NicheMapR R package [Kearney and Porter 2017]. The primary function to call to calculate soil temperature is `soil_temperature()`. The function is a wrapper to solve the differential equations below for equilibrium conditions.  The called function `soil_temperature_function()` contains the majority of the soil temperature model and calls two helper functions.  The two functions implement an equation for calculating soil temperature adapted from Beckman et. al. [1973]; `soil_temperature_integrand()` includes the integrand in the equation and is called by `soil_temperature_equation()`, which represents the full equation.  Below we outline the soil temperature model encapsulated by the functions.

The change is soil temperature ($T_s$) through time $t$ and depth $z$ can be approximated with a differential equation
$$\frac{dT_s}{dt}= \frac{k_s}{\rho_s c_s} \frac{\delta^2T_s}{\delta z^2}$$
where  $k_s$ is the thermal conductivity of soil (units), $\rho_s$ is the density of soil, and $c_s$ is the specific heat of soil (units). We solve the differential equation by breaking the soil profile into depth nodes, exchanging heat. We estimate the steady state temperature profile, which is independent of initial conditions.  The choice of initial conditions only affects the rate of convergence to the steady state solution. A finite difference approximation of the differential equation yields a transient equation for temperature at depth $i$ and time $p$ where we let $a= \frac{k_s}{\rho_s c_s}$:
$$\frac{T_i^{p+1}-T_i^p}{dt}= a \frac{T_{i+dz}^p+T_{i-dz}^p-2T_i^p}{dz^2}. $$
Solving for $T_i^{p+1}$ produces as expression for solving soil temperature at depth $i$ at any time:
$$T_i^{p+1} = a \frac{dt}{dz^2}(T_{i+dz}^p+T_{i-dz}^p)+(1-2a \frac{dt}{dz^2})T_i^p$$

##Soil boundary conditions
Solving the equation requires two boundary conditions: 1) the deep soil temperature, and 2) the soil surface temperature. We assume that the soil temperature at 60cm is constant diurnally. The soil surface temperature at time $p$ is estimated using an energy balance equating the energy conducted to the soil surface the the net heat transfer to the surface by solar radiation, infrared radiation, and conduction [Beckman et al. 1973]:
$$Q_{net_s}= Q_{sol_s} + Q_{emit_s} + Q_{conv_s} +Q_{cond_s}$$
where the solar radiation absorbed by the soil surface is the product of incoming radiation $S$ and the proportion radiation absorbed $\alpha_s$, $Q_{sol_s}= \alpha_s S$. The net thermal radiation to the soil surface is determined by balance of radiation emitted by the soil surface $\sigma T_s^4$ and sky radiation absorbed by the soil surface $\sigma T_{sky}^4$: $Q_{emit_s}= \epsilon_s \sigma [(T^p_{sky})^4-(T^p_0)^4]$, where $\epsilon_s$ is the thermal emissivity of the soil surface and $T^p_{sky}=0.552 T^{1.5}$. Convection between the soil surface and air is estimated as a function of the temperature difference $Q_{conv_s}= h_s(T^p-T^p_s)$, where $h_s$ is the surface heat transfer coefficient. Consideration of temperature profiles and surface roughness yields a expression during neutral conditions [Sellers 1965]: 
$$h_s= \frac{\rho_s c_s k^2 V_r}{ln(z_r/z_o+1)^2},$$
where $V$ is wind speed at reference height $z_r$ and $z_0$ is roughness length. Conduction within the soil profile is estimated as $Q_{cond_s}=k_s \frac{\delta T_s}{\delta z}$, at $z=0$. Approximating $Q_{cond_s}$ by finite difference yields $Q_{cond_s}=\frac{k_s}{dz}(T^p_{dz}-T^p_s)$, at $z=0$

The net heat transfer through the profile across time is expressed as $$Q_{net_s}= \frac{\rho_s c_s dz}{2 dt} (T_0^{p+1}-T_0^p)$$. Letting $a=\frac{2dt}{\rho_s \c_s dz}$ and solving for $T_0^{p+1}$ in the soil energy budget yields the following boundary condition for soil surface temperature:
$$T_0^{p+1}= a \alpha_s S + a \epsilon_S \sigma [(T^p_{sky})^4-(T^p_0)^4]+a h_s (T^p-T_0^p) +a \frac{k_s}{dz}(T^p_{dz}-T^p_0)+T_0^p$$

Soil temperature can be estimated in R as follows:
```{r}
library(zoo)
library(deSolve)

temp_vector= runif(96, min=-10, max=10)
wind_speed_vector= runif(96, min=0, max=0.4)
time_vector= rep(1:24,4)
solrad_vector= rep(c(rep(0,6),seq(10,700,length.out=6), seq(700,10,length.out=6),rep(0,6)),4)

soil_temperature(z_r.intervals=12,z_r=1.5, T_a=temp_vector, u=wind_speed_vector, Tsoil0= 20, z0=0.02, SSA=0.7, TimeIn=time_vector, H= solrad_vector, water_content=0.2, air_pressure=85, rho_so=1620, shade=FALSE)
```

## Soil specific heat and conductivity
We can approximate soil specific heat, $c_s$,as the weighted sum of the specific heats of soil constituents [Campbell and Norman 1998]: $\rho_s c_s= rho_m c_m x_m + rho_o c_o x_o + rho_w c_w x_w,$
where $\rho_s$ is the particle density of soil (bulk density), $x$ is the volume fraction, and the subscripts $m$,$o$,and $w$ stand for minerals, organic matter, and water, respectively. We provide a function to calculate soil specific heat given the volume fractions and $\rho_s$:
```{r}
soil_specific_heat(0.01, 0.6, 0.2, 1620)
```

Various methods of approximating soil thermal conductivity $k_s$ are reviewed in Boguslaw and Lukasz [XXXX]. We provide a function `soil_conductivity()` to estimate $k_s$ using the methods of de Vries [1963]. The function calculates $k_s$ as the weighted average of the thermal conductivities $k$ of soil components $i$:
$$k_s= \frac{\sum j_i k_i x_i}{\sum j_i x_i},$$
where quanity $j_i$ is the ratio of the average temperature gradient in the granules and the corresponding quantity in the medium (water or air):
$$j_i= \frac{1}{3} \sum_{p=a,b,c}[1+\left( \frac{k_i}{k_0}-1 \right)g_{p,i}]^{-1}, $$
where $k_0$ is the conductivity of the continuous medium in which soil particles are suspended (water or air) and $g$ is a shape factor for the soil particles. The soil particles are assumed to be ellipsoids with axes $g_a$, $g_b$, and $g_c$, where $g_a +g_b +g_c=1$.  de Vries [1952] suggests $g_a=g_b=0.125$. We provide the function as follows:
```{r}
soil_conductivity(x=c(0.10,0.40,0.11,0.01,0.2, 0.18), lambda=c(0.10,0.40,0.11,0.01,0.2, 0.18), g_a=0.125)
```

#Example microclimate calculations

```{r}
#LOAD SITE DATA
#setwd("/Volumes/GoogleDrive/My\ Drive/Buckley/Work/Butterflies/Evolution/ClimateData/Niwot") 
#a1.clim= read.csv("A1climate_example.csv")
#site="A1"
lat=40.01; lon=-105.37; elev=2195

airpressures=airpressure_from_elev(elev)

doy=200
Tmn=12
Tmx=33

#surface roughness
z_0_1=0.06

```


##Estimate diurnal variation

```{r}
#estimate daylength
SolarNoon= solar_noon(doy, lon) 
DayLength= daylength(doy, lat)
Ts= SolarNoon+ DayLength/2
Tr= SolarNoon- DayLength/2

# Calculate the diurnal temperature trends
hrs=1:24
Thr= sapply(hrs, FUN=diurnal_temp_variation_sinesqrt, tr=Tr, ts=Ts, T_max=Tmx, T_min=Tmn,T_minp=Tmn) 
Thr1= sapply(hrs, FUN=diurnal_temp_variation_sine, T_max=Tmx, T_min=Tmn)
Thr2= sapply(hrs, FUN=diurnal_temp_variation_sineexp, T_max=Tmx, T_min=Tmn, t_r=Tr, t_s=Ts, alpha=2.59, beta= 1.55, gamma=2.2)

#plot
plot(hrs, Thr, type="l", ylim= range(10,35))
points(hrs, Thr1, col="blue", type="l")
points(hrs, Thr2, col="orange", type="l")
```

##Estimate solar radiation
```{r}
#Estimate solar radiation
taus= rep(0.4,24)        
      
#calculate zenith in radians
psis_deg= sapply(hrs, FUN=zenith_angle, doy=doy, lat=lat, lon=lon)
#convert to radians
psis= degree_to_radian(psis_deg)

#radiation, returns direct, diffuse, and reflected radiation
Qhr= sapply(psis, FUN=estimate_radiation, doy=doy, tau=0.6, elev=1500, rho=0.7)

#Illustrate how to partition radiation by repartitioning
Qhr_tot= colSums(Qhr)
kds= sapply(Qhr_tot, FUN=partition_solar_radiation, method="Erbs", kt=0.6, lat = lat)

Qhr_dir= Qhr_tot*(1-kds)
Qhr_dif= Qhr_tot*(kds)

#plot radiation
plot(hrs, Qhr[1,], type="l" )
points(hrs, Qhr[2,], col="blue", type="l")
points(hrs, Qhr[3,], col="orange", type="l")

points(hrs, Qhr_dir, type="l", lty="dashed")
points(hrs, Qhr_dif, col="blue", type="l", lty="dashed")
```

##Temperature profile

```{r}
      surface_roughness(u_r=c(0.01,0.025,0.05,0.1,0.2), zr=c(0.05,0.25,0.5,0.75,1))
      wind_speed_profile_neutral(u_r=0.1, zr=0.1, z0=0.2, z=0.15)
      air_temp_profile_neutral(T_r=20, zr=0.1, z0=0.2, z=0.15, T_s=25)
    
      wind_speed_profile(u_r=0.1, zr=0.1, z0=0.2, z=0.15)
      air_temp_profile(T_r=20, u_r=0.1, zr=0.1, z0=0.2, z=0.15, T_s=25)
    
```

##Degree days

```{r}
DD1= degree_days(T_min=Tmn, T_max=Tmx, LDT=12, UDT=33, method="single.sine")
DD2= degree_days(T_min=Tmn, T_max=Tmx, LDT=12, UDT=33, method="single.triangulation")
```

##Soil temperatures

```{r}
#Calculate soil temperature
#' temp_vector= runif(96, min=-10, max=10)
#' wind_speed_vector= runif(24, min=0, max=0.4)
#' time_vector= rep(1:24,4)
#' solrad_vector= rep(c(rep(0,6),seq(10,700,length.out=6), seq(700,10,length.out=6),rep(0,6)),4)
#'

#Simulate wind speed
Vs= runif(24, min=0, max=0.4)

#Estimate surface roughness
z0=0.02

#require in package?
library(zoo)
library(deSolve)

Tsoil=soil_temperature(z_r.intervals=12,z_r=1.5, T_a=Thr, u=Vs, Tsoil0= 20, z0=z0, SSA=0.7, TimeIn=hrs, H= Qhr_tot, water_content=0.2, air_pressure=airpressures, rho_so=1620, shade=FALSE)
#Tsoil=soil_temperature_noint(z_r.intervals=12,z_r=1.5, T_a=Thr, u=Vs, Tsoil0= 20, z0=z0, SSA=0.7, TimeIn=hrs, H= Qhr_tot, water_content=0.2, p=airpressures, rho_so=1620, shade=FALSE)

#plot
#plot(hrs, Tsoil, type="l")
```


  






