---
title: "Estimating microclimates"
author: "Lauren Buckley"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
##Load TrenchR package
library(TrenchR)
```

##Solar angles and daylength
Estimates of sun position and day length are often estimated based on calendar date. We provide a function `day_of_year()` that facilitates translating a date into a calendar date. We next present functions to calculate sun angles.  We use the calendar date (J) to estimate solar declination (radian), which is the angular distance of the sun north or south of the earth’s equator, as follows (`dec_angle()`):
$$ \delta= \arcsin[0.39795 \cos(0.21631+ 2 \arctan (0.967 \tan [0.0086(-186+J)]))].$$

Zenith angle ($\psi$, degrees), the location of the sun as an angle measured from vertical, can be estimated as:
$$\cos \psi = \sin( \delta) \sin( \phi) + \cos(\delta)\cos(\phi)\cos(\pi/12*(h-h_0)),$$
where J is calendar date, $\phi$ is latitude (radians), h is hour, and $h_0$ is the time of solar noon. The time of solar noon (`solar_noon()`) can be estimated as $h_0= 12 - LC -ET$, where LC in the longitude correction (+4 minutes for each degree east of the standard meridian) and ET is the equation of time.  The ET is estimated as: 
$$ ET= \frac{-104.7\sin (f)+596.2\sin (2f)+4.3\sin (3f)-12.7\sin (4f)-429.3\cos (f)-2.0\cos (2f)+19.3\cos (3f)}{3600}, $$
where $f=279.575 + 0.9856 J$, in degrees. 

The azimuth angle, the angle (in degrees) from which the sunlight is coming measured from true north or south measured in the horizontal plane. The azimuth angle ($AZ$ in degrees) is estimated with respect to due south, increasing in the counter clockwise directi such that $90^\circ$ is east. $AZ$ can be calculated as a function of declination angle $\delta$, zenith angle $\psi$ and latitude $\phi$ as follows:
$$\cos AZ= \frac{-(\sin \delta - \cos \psi \sin \phi)}{\cos \phi \sin \psi}.$$

These sun angles can be used to solve for daylength. The equation for half daylength $h_s$, which is the time (in degrees) from sunrise to solar noon, is used as the basis for our `daylength()` function:
$$ h_s= \cos^{-1}\left( \frac{\cos \psi-sin \phi \sin \delta}{\cos \phi \cos \delta} \right).$$

We provide these R functions as follows: 
```{r}
day_of_year("2017-04-22", format= "%Y-%m-%d")
dec_angle(112)
zenith_angle(doy=112, lat=47.61, lon=-122.33, hour=12)
solar_noon(lon=-122.335, doy=112)
azimuth_angle(doy=112, lat=47.61, lon=-122.33, hour=12)
daylength(47.61, 112)
```

## Radiation estimates
We provide functions to estimate solar (shortwave) radiation by discounting incoming solar radiation due to scattering as the radiation travels through the atmosphere. The primary function `estimate_radiation()` returns estimates of the direct, diffuse, and reflected components of solar radiation ($W m^{-2}$). The radiation streams are the direct irradiance on a surface perpendicular to the beam $S_p$, the diffuse sky irradiance on a horizontal plane $S_d$ (W m−2), and the reflected radiation from the ground $S_r$. The sum of these three streams represents the total irradiance of a horizontal surface $S_t$.

Calculation of these flux densities requires the introduction of several additional quantities. The atmospheric transmissivity, $\tau$, ranges between 0.6 and 0.7 for typical clear sky conditions (Gates 1980). The solar constant, $S_{p0}$, indicates extraterrestrial flux density to be 1,360 $W m^{−2}$. Optical air mass number, $m_a$, is the ratio of slant path length through the atmosphere to zenith path length and is a function of atmospheric pressure: $m_a=p_a/(101.3\cos(\psi))$, where $p_a$ is air pressure (kPa) and $\psi$ is zenith angle (see above)$. The function `airpressure_from_elev()` estimates air pressure as $p_a (kPa)= 101.3 \exp \frac{-E}{8200}$, where $E$ is the elevation in meters above sea level. 

Direct irradiance is a function of the distance a solar beam travels through the atmosphere; the transmittance of the atmosphere, $\tau$; and the incident flux density, $\S_{p0}$: $S_p=S_{p0}\tau^{m_a}$. We approximate diffuse radiation using an empirical relation [Liu and Jordan 1960], $S_d=0.3(1-\tau^{m_a}) S_{p0}\cos\psi$.Finally, reflected radiation is the product of albedo, $\rho_S$ and the sum of direct and diffuse radiation: $S_r= \rho_S (S_p+S_d)$.

We additionally provide a function (`monthly_solar_radiation()`) implementing an algorithm to estimate average monthly solar radiation ($W m^{-2}$ per day) using basic topographic and climatic information as input.

We provide a function `partition_solar_radiation()` to estimate the diffuse fraction $k_d$ that can be used to partition measures of solar radiation into direct and diffuse components.  The partitioning is based on empirically-derived relations estimated as a function of the clearness index $k_t$, which is the ratio of the global solar radiation measured at the surface to the total solar radiation at the top of the atmosphere. The relations are compiled in Wong and Chow [2001, Applied Energy 69:1991-224].

The functions are available in the TrenchR package as follows:
```{r}
airpressure_from_elev(1500)
estimate_radiation(doy=112, psi=1, tau=0.6, elev=1500, rho=0.7)
monthly_solar_radiation(lat=47.61,lon=-122.33,doy=112,elev=1500,Temp=15,Hr=50,P=50)
partition_solar_radiation(500, 0.5)
```

## Temperature and windspeed profiles
We scale Ta across height by estimating a wind speed profile using the relationship:
$$u(z)= \frac{u^*}{0.4}ln \frac{z-d}{z_m} $$
where u(z) is the wind speed (m/s) at height z (m), u* is the friction velocity (m/s), d is the zero plane displacement (m), and the constant 0.4 is the von Karman constant.    We fit this relationship to data collected at heights of 0.05, 0.25, 0.5, 0.75, 1, 1.25, 1.5m at 1 minute intervals for 12 days in July 2013 at the subalpine site (C1).  We estimated d as the y-intercept of the relationship between u and ln(z). We then estimated surface roughness, z0, as exp(-b/n) from fitting the relationship u=b+n ln(z-d), where u is wind velocity (m/s) and z is measurement height. 

We used the following relationship to estimate air temperature and wind speed at height z [2,3]:
$$\frac{u_z}{u_r}=\frac{ln(z/z_0+1)}{ln(z_r/z_0+1)}=\frac{T_z-T_0}{T_r-T_0} $$
where uz and ur are the wind speeds (m/s) at height z (m) and reference height r (m), respectively.  Tz, Tr, and T0 are air and soil temperatures at height z and r and soil surface temperature, respectively. 

```{r}
#surface_roughness
air_temp_profile(T_r=20, u_r=0.1, zr=0.1, z0=0.2, z=0.15, T_s=25)
air_temp_profile_neutral(T_r=20, zr=0.1, z0=0.2, z=0.15, T_s=25)
```

```{r}
#wind_speed_profile
#wind_speed_profile_neutral
#wind_speed_profile_segment_in_beta
```

## Diurnal variation in temperature and radiation

```{r}
diurnal_temp_variation_sine() #fix parameters
diurnal_temp_variation_sineexp(Tmx, Tmn, Hr, tr, ts, alpha = 2.59,
  beta = 1.55, gamma = 2.2) #fix
diurnal_temp_variation_sinesqrt( Hr=8, Hn=6, Ho=18, Tmx=30, Tmn=20, Tmn_p=22)
```

```{r}
diurnal_radiation_variation(doy=112, solrad=500, hour=12, lon=-122.33, lat=47.61)
```

## Degree days
```{r}
degree_days(7,14,12,33,"single.sine")
```

## Soil temperature
Ta and radiation were used as inputs for a microclimate model [2,6,7], which used finite-difference methods to solve heat balance equations describing soil temperatures at the surface and specified depths [8,9]. 

We implemented a microclimate model [31–33] using finite-difference methods to solve heat balance equations describing soil temperatures at the surface and specified depths [17,34]; predicted surface temperature is then used to compute air temperature profiles above the surface.



```{r}
#soil_conductivity
#soil_specific_heat
#soil_temperature
#soil_temperature_noint
#soil_temp_integrand
#soil_temp_overall_function
#Tsoil
```


