---
title: "Biophysical model usecase"
author: "Aji John"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


This vignette serves as a usecase bringing together various products under the Trench Project umberalla to calculate Operative Temperature of a lizard. Furthermore, we will transistion into depicting if the organism would be subjected to stress. The flow will start with downloading microclimate data using the API for an area of interest. We will download variables namely GLW(downward flux of near-infrared radiation),SWDOWN(Downward flux of visible radiation),TAH(Air Temp in Canopy),Tair, and Tsurface for a time frame. Once the microclimate data is ready, we will download it, and do some data wrangling. As we are downloading the hourly mean temperatures by date, we will be using only the noon mean temperature. We will be using Fei et. al model to calculate the body temperature.
[1] T. Fei, A. K. Skidmore, V. Venus, T. Wang, M. Schlerf, B. Toxopeus, S. van Overjijk, M. Bian, and Y. Liu, “A body temperature model for lizards as estimated from the thermal environment,” J. Therm. Biol., vol. 37, no. 1, pp. 56–64, 2012.

## Install the packages

```{r, echo=FALSE, results='asis'}
library(devtools)
library(tidyverse)
devtools::install_github("trenchproject/TrenchR")
library(TrenchR)
```
Assumes 'devtools' is installed, if not, you can install via CRAN.

## Loading microclimate data

```{r, echo=FALSE, results='asis'}
SampleMicroclimate <- read_delim("./SampleMicroclimate.csv", 
    "\t", escape_double = FALSE, col_types = cols(GLW = col_number(), 
         SWDOWN = col_number(), TAH = col_number(), 
        Tair = col_number(), To_Lizard = col_number(), 
        Tsurface = col_number(), datetime = col_number(), 
       datetimef = col_datetime(format = "%Y-%m-%d %H:%M:%S"), 
        day = col_number(), hr = col_number(), 
       lat = col_number(), lon = col_number(), 
        month = col_number(), year = col_number()), 
    trim_ws = TRUE)
```

## Runing a biophysical model [Fei et al]

The figure sizes have been customised so that you can easily put two images side-by-side. 

```{r, fig.show='hold'}
#Now, lets invoke the Fei et. al model to calculate the body temperature
#calculate_To(TAH, Tsurface, Tair, SWDOWN, GLW, shade)
#GLW(downward flux of near-infrared radiation)
#SWDOWN(Downward flux of visible radition)
#TAH(Air Temp in Canopy)
#Tair - Air Temperature
#Tsurface - Surface Temperature 
SampleMicroclimate$To_Lizard <- Tb_Fei(SampleMicroclimate$TAH,
                                              SampleMicroclimate$Tsurface,
                                              SampleMicroclimate$Tair,
                                              SampleMicroclimate$SWDOWN,
                                              SampleMicroclimate$GLW,
                                              0)
```


## Plot the distribution of Body temperature using GGPlot



```{r, echo=FALSE, results='asis'}
#lets see the spread
SampleMicroclimate %>% ggplot(aes(To_Lizard)) + geom_histogram(bins = 15)
```
